\MakeUppercase{Graph Product Structure for Non-Minor-Closed Classes}

### 2022-03-16 response.tex

\begin{itemize}
\item graphs drawn on a fixed surface with each edge in $\leq k$ crossings,
\item map graphs of degree $\leq k$ graphs embedded in a fixed surface,
\item string graphs with each curve in $\leq k$ intersections,
\item $k$-nearest neighbour graphs in the plane, and
\item powers of classes of graphs of degree $\leq k$ which admit a product structure theorem.
\end{itemize}

All but the last bullet point are shown to follow from the first.
The main results in the paper are correct, strong, and of high interest,
and the paper should definitely be accepted to JCTB after some
(fairly significant) revisions. In particular, the proof of Theorem 3 is
difficult to read due to the notation, and the paper could be substantially
streamlined without losing the above results. Detailed suggestions follow.

Major comments:

The following comments (and several additional minor comments) are all related to the proof of Theorem~3 in the original submission.  Much of the difficulty comes from the fact that there are many things to keep track of. We are a given a graph $G$, a shortcut system $\mathcal{S}$, a layering $\mathcal{L}$, a vertex-partition $\mathcal{P}$, and a tree decomposition $\mathcal{T}$ of the quotient graph $H:=G/\mathcal{P}$.  Now we have to use these to show the existence of a new layering $\mathcal{L}'$, a new vertex partition $\mathcal{P'}$, a new tree decomposition $\mathcal{T}'$ of the new quotient graph $H':=G^{\mathcal{S}}$.

@@

During the proof, a number of related sets come up, which we have given names to. As a referee point out, these could remain unnamed and be described using an appropriate notation.

\todo[inline]{Write something more here depending on how the experiment with new notation works out.}

One issue is that a number of the proofs in this section are essentially "definition checking".  The relationships between the various vertex subsets that arise in the proof really do just follow (by design) from their definitions.  In the paper, we do this checking explicitly, though each of these proofs could also be replaced by a sentence of the form ``This [claim/lemma] follows from the definitions of\ldots''.

However, when discussing the proof of Theorem~3 with colleagues (very familiar with product structure) recently, we found them unconvinced by high-level explanations.  It was only after carefully unpacking the definitions that they became convinced.  These conversations usually end with someone saying ``Ah, of course!''

### 2022-03-16 k-planar.tex
Noticed a missing definition, to fix later
Much simpler.....

We begin with a standard technique that allows us to work with a normalised tree-decomposition:

@@

%\subsection{Generalized Tripod Partitions}

\begin{figure}[htbp]
  \begin{center}
    \includegraphics{figs/tripoddo}
  \end{center}
  \caption{The sets $Y_x$, $F_x$, and $V_x$ associated with $x\in V(T)$
  and the ancestors $a_1,\ldots,a_{t'}$ of $X$ such that $F_x \subseteq \bigcup_{i=1}^{t'} Y_{a_i}$.}
  \label{fig:generalized-tripod}
\end{figure}

The following lemma shows how to interpret an $H$-partition of $G$ and a tree-decomposition of $H$ as a `hierarchical' decomposition of $G$; refer to \cref{fig:generalized-tripod}.

\begin{lem}\label{generalized-tripod}
  Let $G$ be a graph; let $\mathcal{L}:=\langle L_1,\ldots,L_h\rangle$ be a layering of $G$; let $\mathcal{Y}:=(Y_x: x\in V(H))$ be an $H$-partition of $G$ of layered width at most $\ell$ with respect to $\mathcal{L}$ where $H$ has treewidth at most $t$; and let $\mathcal{T}:=(B_x:x\in V(T))$ be a tree-decomposition of $H$ satisfying the conditions of \cref{nice-decomposition}.  For each $x\in V(T)$, let $V_x := \bigcup_{y\in V(T_x)} Y_y$, $F_x:=\{w\in V(G): vw\in E(G), v\in V_x,\, w\not\in V_x\}$, and $N_x:=V_x\cup F_x$.  Then,
  \begin{compactenum}[(Y1)]
    % \item\ylabel{thickness} $\mathcal{Y}=(Y_x: x\in V(T))$ is a partition of $V(G)$ of layered width at most $\ell$ with respect to $\mathcal{L}$.
    \item\ylabel{separator} For each $x\in V(T)$, there is no edge $vw\in E(G)$ with $v\in V_x$ and $w\in V(G)\setminus N_x$.
    \item\ylabel{ancestor-edge} For each $x\in V(T)$, there is a set $\{a_1,\ldots,a_{t'}\}$ of $t'\le t$ strict $T$-ancestors of $x$ such that $F_x \subseteq \bigcup_{i=1}^{t'} Y_{a_i}$.
  \end{compactenum}
\end{lem}

\referee{1}{1. Lemma 3: This lemma is really about ``normalised'' tree-decompositions, and says that for each $x \in V(T)$, the set $V(T_x)$ has at most $t$ neighbours in $H$. Call this property (T3) and remove (Y1) through (Y5), which should now be obvious.}

\note{PM}{I've added to the \texttt{response.tex} file explaining some of this.  Each of (Y1), (Y3), (Y4), and (Y5) have a one-sentence justification. The justification for (Y2) is just definition unpacking and then using the observation that the referee recommends.}

\note{PM}{I plan to do an experiment with the referee's suggested notation, in a separate git branch to see how it turns out.}

Before proving \cref{generalized-tripod} we point out more properties that are immediately implied by it:

\begin{compactenum}[(Y1)]\setcounter{enumi}{2}
  \item\ylabel{y-subsets} $Y_x\subseteq V_x$ for every $x\in V(T)$.
  \item\ylabel{containment-i} $V_x\subseteq V_a$ for every $T$-ancestor $a$ of $x$.
  \item\ylabel{containment-ii}$N_x\subseteq N_a$ for every $T$-ancestor $a$ of $x$.
\end{compactenum}

Property~\yref{y-subsets} follows from the fact that $V_x$ is the union of several sets, one of which is $Y_x$.  Property~\yref{containment-i} follows from the definition of $V_x$ and the fact that $V(T_x)\subseteq V(T_a)$. To show Property~\yref{containment-ii} first note that, by \yref{containment-i} it suffices to consider vertices $w\in F_x=N_x\setminus V_x$. By definition, every vertex $w\in F_x$ is adjacent, in $G$, to a vertex $v\in V_x$.  By \yref{containment-i}, $v\in V_a$, so $w$ is either in $V_a$ or $w$ satisfies the condition $vw\in E(G)$, $v\in V_a$, and $w\not\in V_a$, so $w\in F_a$.  In either case $w\in N_a=V_a\cup F_a$.  Note that none of \yref{y-subsets}--\yref{containment-ii} depends on \yref{ancestor-edge} (which is important, since \yref{containment-i} is used to establish \yref{ancestor-edge} in the following proof).

\begin{proof}[Proof of \cref{generalized-tripod}]
  % Property~\yref{thickness} follows immediately from the fact that $V(T)=V(H)$  and the fact that $\mathcal{Y}$ has layered width at most $\ell$ with respect to $\mathcal{L}$.
  Property \yref{separator} is immediate from the definitions of $F_x$ and $N_x$.  In particular, $(N_x,V(G)\setminus V_x)$ is a separation of $G$ with $F_x=N_x\cap(V(G)\setminus V_x)$.

  To establish Property~\yref{ancestor-edge}, consider some vertex $w\in F_x$.  Since $w\in F_x$, there exists an edge $vw\in E(G)$ with $v\in V_x$ and $w\not\in V_x$.  Since $v\in V_x$, $v\in Y_{x'}$ for some $T$-descendant $x'$ of $x$ (possibly $x=x'$). Since $\mathcal{Y}$ is a partition, $w\in Y_{a}$ for some $a\not\in V(T_x)$.  Since $vw\in E(G)$, we have $x'a\in E(H)$.  By \tref{ancestor-edge}, one of $a$ or $x'$ is a $T$-ancestor of the other. Since $w\in Y_a\subseteq V_a$ and $w\not\in V_x\supseteq V_{x'}$, \yref{containment-i} rules out the possibility that $x'$ is a $T$-ancestor of $a$. Therefore, $a$ is a $T$-ancestor of $x$ which is a $T$-ancestor of $x'$.  Let $z_0,\ldots,z_r$ be the path in $T$ from $z_0:=x'$ to $z_r:=a$.  Since $x'a\in E(H)$, at least one of $a$ or $x'$ is in $B_{z_i}$, for each $i\in\{0,\ldots,r\}$ and there is at least one $i$ such that $B_{z_i}$ contains both $x'$ and $a$.  However, by \tref{subtree-root} $x'$ is not contained in $B_{z_i}$ for any $i\in\{1,\ldots,r\}$.  Therefore $a\in B_{z_i}$ for each $i\in\{0,\ldots,r\}$.  In particular, $a$ is contained in $B_x$.
  Property~\yref{ancestor-edge} now follows from the fact that $|B_x|\le t+1$ and $B_x$ contains $x$.
\end{proof}

@@

\referee{1}{2. Proof of Theorem 9: There is too much notation throughout. As
	one example, it is hard to remember which of $T_x$, $Y_x$, $V_x$, $F_x$, $N_x$,
	$S_x$, $X_v$, $B_x$, $C_x$ is which. If $\phi(S)$ is written for the subset of $V(G)$
	corresponding to a set $S \subseteq V(H)$ and appropriate shorthand and
	standard notation for neighbourhoods are used, then $Y_x$, $V_x$, $F_x$, and
	$N_x$ can be replaced by $\phi(x)$, $\phi(T_x)$, $N(\phi(T_x))$, and $V(G) \setminus N[\phi(T_x)]$.
	The set $X_v$ is only used in the statement of Claim~1, so it does not
	need a name; say what these vertices are explicitly. It may be more
	indicative to write $\phi'(x)$ for $S_x$ and $B'_x$ for $C_x$.}

\begin{proof}
Apply \cref{generalized-tripod} to $G$ and let $\mathcal{L}$, $\mathcal{Y}$, $\mathcal{T}$, $T$, $Y_x$, $V_x$, $F_x$, and $N_x$ be defined as in \cref{generalized-tripod}, where the partition $\mathcal{Y}$ has width $\ell$ with respect to the layering $\mathcal{L}$.

For a node $x\in V(T)$, we say that a shortcut $P\in\SS$ \defin{crosses} $x$ if $Y_x$ contains an internal vertex of $P$, that is, $P=(v_0,\ldots,v_r)$ and $\{v_1,\ldots,v_{r-1}\}\cap Y_x\neq\emptyset$.  We say that a vertex $v\in V(G)$ \defin{participates} in $x$ if $v\in Y_x$, or $\SS$ contains a shortcut $P$ with $v\in V(P)$ and $P$ crosses $x$. We let $X_v$ denote the set of nodes $x\in V(T)$ such that $v$ participates in $x$.

\begin{clm}\label{x-v-ancestor}
  For any $v\in V(G)$ there exists a (unique) node $a(v)\in X_v$ such that
  $a(v)$ is a $T$-ancestor of every node in $X_v$.
\end{clm}

\begin{proof}
  Let $Z := \{v\} \cup \{\{v_1,\ldots,v_{r-1}\}:(v_0,\ldots,v_r)\in\SS, v\in \{v_0,\ldots,v_r\}\}$. Then $G[Z]$ is connected because $Z$ is the union of (vertex sets of) paths in $G$, each of which contains $v$.

  We claim that $v$ participates in a node $x\in V(T)$ if and only if $Z\cap Y_x\neq\emptyset$.  If $v$ participates in $x$ then either $v\in Y_x$, so $Z\cap Y_x\supseteq\{v\}$; or $v\in \{v_0,\ldots,v_r\}$ for some shortcut $(v_0,\ldots,v_r)\in\SS$ that crosses $x$, so $Z\cap Y_x\supseteq \{v_i\}$ for some $i\in\{1,\ldots,r-1\}$.  In the other direction, if $Z\cap Y_x\neq\emptyset$, then either $Z\cap Y_x\supseteq \{v\}$, so $v\in Y_x$; or $Z\cap Y_x\supseteq \{v_i\}$ where $i\in\{1,\ldots,r\}$, $(v_0,\ldots,v_r)\in\SS$ and $v\in\{v_0,\ldots,v_r\}$, so $v\in V(P)$ for a path $P=(v_0,\ldots,v_r)\in\SS$ that crosses $x$.

  Let $X_H:=\{x\in V(H): Z\cap Y_x\neq\emptyset\}$.  The connectivity of $G[Z]$ implies that $H[X_H]$ is connected.
  %Property~\tref{ancestor-edge} implies that, for every $xy\in E(H)$, one of $x$ or $y$ is a $T$-ancestor of the other.
  Choose $a(v)\in X_H$ to be the member of $X_H$ that does not have a strict $T$-ancestor in $X_H$.  Transitivity of the $T$-ancestor relationship, \tref{ancestor-edge}, and connectivity of $H[X_H]$ implies that such an $a(v)$ exists and is a $T$-ancestor of every node $x\in X_H$, as required.
\end{proof}

@@

For each $x\in V(T)$, define $S_x := \{v\in V(G): a(v)= x\}$. Observe that $\PP:=(S_x : x\in V(T))$ is a partition of $V(G)$.\footnote{The sets in $\PP$ are disjoint because $a:V(G)\to V(T)$ is a function.  The sets in $\PP$ cover $V(G)$ since, for each $v\in V(G)$, $v\in S_{a(v)}$.} Let $J:=G^\SS/\PP$ denote the resulting quotient graph. We consider $V(J)\subseteq V(T)$, where each $x\in V(J)$ is the vertex obtained by contracting $S_x$ in $G^{\SS}$. (Any node $x\in V(T)$ with $S_x=\emptyset$ does not contribute a vertex to $J$.)

@@

First we need to understand the relationship between sets $S_x\in\PP$ and the related sets $Y_x$, $V_x$, $F_x$, and $N_x$.

\referee{1}{8. general comment: Claims 1, 2, and 4 are especially belaboured by
notation as they are quite straightforward.}

\note{PM}{Same comment: Yes, these claims are definition checking.  We'll see if the experimental notation change helps.}

\begin{clm}
	\label{s-subset}
  For every $x\in V(T)$, $S_x\subseteq V_x$.
\end{clm}

\begin{proof}
  For the sake of contradiction, assume otherwise, so there exists some $v\in S_x\setminus V_x$. By \yref{y-subsets}, $Y_x\subseteq V_x$, so $v\not\in Y_x$.  Therefore, $\SS$ contains a path $P$, with $v\in V(P)$, that crosses $x$.  The path $P$ contains a proper subpath $v_0,v_1,\ldots,v_{r}$ such that $v=v_0$ and $v_r\in Y_x$. Since $v\not\in V_x$ and $v_r\in Y_x\subseteq V_x$, \yref{separator}, implies that $v_i\in F_x$ for some $i\in\{0,\ldots,r-1\}$. Now \yref{ancestor-edge} implies $v_i\in Y_a$ for some strict $T$-ancestor $a$ of $x$.  Therefore, either $v\in Y_a$ or $P$ crosses $a$. But this implies that $a(v)$ is a $T$-ancestor of $a$, which is a strict $T$-ancestor of $x$, contradicting the assumption that $v\in S_x$.
\end{proof}

Next we complete Step~(i) and show that $\PP$ has small layered width with respect to the layering $\mathcal{L}=\langle L_1,\ldots,L_h\rangle$:

@@

\referee{1}{10. page 10, Claim 5: Indices could mostly be removed by continuing
to talk about $T$-ancestors instead. Move the definition of $H+$ and its
directed version up, along with the explanation of what will be proven,
so that it isn’t necessary to define the $s_i$.}

\note{PM}{I also found the steps in this proof tedious. Unfortunately, when I tried to give a simple explanation for it in Banff, both Piotr and Torsten remained unconvinced until we opened the paper and walked through the steps of the proof together.  I suspect the referee who can see through these proofs if one of the authors of \cite{PS21}.}

>>>>

We begin with a standard technique that allows us to work with a normalised tree-decomposition whose tree has same vertex set as the graph it decomposes:

@@

%\subsection{Generalized Tripod Partitions}
% 
% 
% \begin{figure}[htbp]
%   \begin{center}
%     \includegraphics{figs/tripoddo}
%   \end{center}
%   \caption{The sets $Y_x$, $F_x$, and $V_x$ associated with $x\in V(T)$
%   and the ancestors $a_1,\ldots,a_{t'}$ of $X$ such that $F_x \subseteq \bigcup_{i=1}^{t'} Y_{a_i}$.}
%   \label{fig:generalized-tripod}
% \end{figure}

% The following lemma shows how to interpret an $H$-partition of $G$ and a tree-decomposition of $H$ as a `hierarchical' decomposition of $G$; refer to \cref{fig:generalized-tripod}.

% For a vertex $v$ of $G$, $N_G(v):=\{w\in V(G):vw\in E(G)\}$ and $N_G[v]:=N_G(v)\cup\{v\}$.  For a vertex set $S\subseteq V$, $N_G(S):=\bigcup_{v\in S} N_G(S)\setminus S$ and $N_G[S]:=N_G(S)\cup S$.

% \note{PM}{
% Guide to old notation:
% \begin{compactitem}
%   \item $V_x := \bigcup_{y\in V(T_x)} Y_y = G_x$
%   \item $F_x:=\{w\in V(G): vw\in E(G), v\in V_x,\, w\not\in V_x\}= N_G(G_x)$
%   \item $N_x:=V_x\cup F_x=N_G[G_x]$
% \end{compactitem}
% }
% 
% \begin{lem}\label{generalized-tripod}
%   Let $G$ be a graph; let $\mathcal{L}:=\langle L_1,\ldots,L_h\rangle$ be a layering of $G$; let $\mathcal{Y}:=(Y_x: x\in V(H))$ be an $H$-partition of $G$ of layered width at most $\ell$ with respect to $\mathcal{L}$ where $H$ has treewidth at most $t$; and let $\mathcal{T}:=(B_x:x\in V(T))$ be a tree-decomposition of $H$ satisfying the conditions of \cref{nice-decomposition}. 
%   For each $x\in V(T)$, let $H_x:=V(T_x)$ and let $G_x:=\bigcup_{y\in H_x}Y_y$.
% 
%   \begin{compactenum}[(Y1)]
%     % \item\ylabel{thickness} $\mathcal{Y}=(Y_x: x\in V(T))$ is a partition of $V(G)$ of layered width at most $\ell$ with respect to $\mathcal{L}$.
%     % \item\ylabel{separator} For each $x\in V(T)$, $N_G(\phi(T_x))$ separates $\phi(T_x)$ from $N
%     % there is no edge $vw\in E(G)$ with $v\in V_x$ and $w\in V(G)\setminus N_x$.
% 
%     \item\ylabel{ancestor-edge} For each $x\in V(T)$, there is a set $\{a_1,\ldots,a_{t'}\}$ of $t'\le t$ strict $T$-ancestors of $x$ such that $N_G(G_x) \subseteq \bigcup_{i=1}^{t'} Y_{a_i}$.
%   \end{compactenum}
% \end{lem}
% 
% \begin{proof}
%   Let $\{a_1,\ldots,a_{t'}\}:= B_x\setminus\{x\}$.  Then, by the definition of a normalized tree decomposition $N_H(H_x))\subseteq\{a_1,\ldots,a_{t'}\}$.  Therefore $N_G(G_x)\subseteq \bigcup_{i=1}^{t'} Y_{a_i}$.
% \end{proof}

% \referee{1}{1. Lemma 3: This lemma is really about ``normalised'' tree-decompositions, and says that for each $x \in V(T)$, the set $V(T_x)$ has at most $t$ neighbours in $H$. Call this property (T3) and remove (Y1) through (Y5), which should now be obvious.}
% 
% \note{PM}{I've added to the \texttt{response.tex} file explaining some of this.  Each of (Y1), (Y3), (Y4), and (Y5) have a one-sentence justification. The justification for (Y2) is just definition unpacking and then using the observation that the referee recommends.}
% 
% \note{PM}{I plan to do an experiment with the referee's suggested notation, in a separate git branch to see how it turns out.}
% 
% Before proving \cref{generalized-tripod} we point out more properties that are immediately implied by it:
% 
% \begin{compactenum}[(Y1)]\setcounter{enumi}{2}
%   \item\ylabel{y-subsets} $Y_x\subseteq V_x$ for every $x\in V(T)$. \note{PM}{$Y_x$, for $x\in V(T)$ is not defined!}
%   \item\ylabel{containment-i} $V_x\subseteq V_a$ for every $T$-ancestor $a$ of $x$.
%   \item\ylabel{containment-ii}$N_x\subseteq N_a$ for every $T$-ancestor $a$ of $x$.
% \end{compactenum}

% Property~\yref{y-subsets} follows from the fact that $V_x$ is the union of several sets, one of which is $Y_x$.  Property~\yref{containment-i} follows from the definition of $V_x$ and the fact that $V(T_x)\subseteq V(T_a)$. To show Property~\yref{containment-ii} first note that, by \yref{containment-i} it suffices to consider vertices $w\in F_x=N_x\setminus V_x$. By definition, every vertex $w\in F_x$ is adjacent, in $G$, to a vertex $v\in V_x$.  By \yref{containment-i}, $v\in V_a$, so $w$ is either in $V_a$ or $w$ satisfies the condition $vw\in E(G)$, $v\in V_a$, and $w\not\in V_a$, so $w\in F_a$.  In either case $w\in N_a=V_a\cup F_a$.  Note that none of \yref{y-subsets}--\yref{containment-ii} depends on \yref{ancestor-edge} (which is important, since \yref{containment-i} is used to establish \yref{ancestor-edge} in the following proof).
% 
% 
% \begin{proof}[Proof of \cref{generalized-tripod}]
%   % Property~\yref{thickness} follows immediately from the fact that $V(T)=V(H)$  and the fact that $\mathcal{Y}$ has layered width at most $\ell$ with respect to $\mathcal{L}$.
%   Property \yref{separator} is immediate from the definitions of $F_x$ and $N_x$.  In particular, $(N_x,V(G)\setminus V_x)$ is a separation of $G$ with $F_x=N_x\cap(V(G)\setminus V_x)$.
% 
%   To establish Property~\yref{ancestor-edge}, consider some vertex $w\in F_x$.  Since $w\in F_x$, there exists an edge $vw\in E(G)$ with $v\in V_x$ and $w\not\in V_x$.  Since $v\in V_x$, $v\in Y_{x'}$ for some $T$-descendant $x'$ of $x$ (possibly $x=x'$). Since $\mathcal{Y}$ is a partition, $w\in Y_{a}$ for some $a\not\in V(T_x)$.  Since $vw\in E(G)$, we have $x'a\in E(H)$.  By \tref{ancestor-edge}, one of $a$ or $x'$ is a $T$-ancestor of the other. Since $w\in Y_a\subseteq V_a$ and $w\not\in V_x\supseteq V_{x'}$, \yref{containment-i} rules out the possibility that $x'$ is a $T$-ancestor of $a$. Therefore, $a$ is a $T$-ancestor of $x$ which is a $T$-ancestor of $x'$.  Let $z_0,\ldots,z_r$ be the path in $T$ from $z_0:=x'$ to $z_r:=a$.  Since $x'a\in E(H)$, at least one of $a$ or $x'$ is in $B_{z_i}$, for each $i\in\{0,\ldots,r\}$ and there is at least one $i$ such that $B_{z_i}$ contains both $x'$ and $a$.  However, by \tref{subtree-root} $x'$ is not contained in $B_{z_i}$ for any $i\in\{1,\ldots,r\}$.  Therefore $a\in B_{z_i}$ for each $i\in\{0,\ldots,r\}$.  In particular, $a$ is contained in $B_x$.
%   Property~\yref{ancestor-edge} now follows from the fact that $|B_x|\le t+1$ and $B_x$ contains $x$.
% \end{proof}

@@

% \referee{1}{2. Proof of Theorem 9: There is too much notation throughout. As
% 	one example, it is hard to remember which of $T_x$, $Y_x$, $V_x$, $F_x$, $N_x$,
% 	$S_x$, $X_v$, $B_x$, $C_x$ is which. If $\phi(S)$ is written for the subset of $V(G)$
% 	corresponding to a set $S \subseteq V(H)$ and appropriate shorthand and
% 	standard notation for neighbourhoods are used, then $Y_x$, $V_x$, $F_x$, and
% 	$N_x$ can be replaced by $\phi(x)$, $\phi(T_x)$, $N(\phi(T_x))$, and $V(G) \setminus N[\phi(T_x)]$.
% 	The set $X_v$ is only used in the statement of Claim~1, so it does not
% 	need a name; say what these vertices are explicitly. It may be more
% 	indicative to write $\phi'(x)$ for $S_x$ and $B'_x$ for $C_x$.}

\begin{proof}
Let $\mathcal{L}:=\langle L_1,\ldots,L_h\rangle$ be a layering of $G$; let $\mathcal{Y}:=(Y_x: x\in V(H))$ be an $H$-partition of $G$ of layered width at most $\ell$ with respect to $\mathcal{L}$; and let $\mathcal{T}:=(B_x:x\in V(T))$ be a tree-decomposition of $H$ satisfying the conditions of \cref{nice-decomposition}.   
% For each $x\in V(T)$, let $H_x:=V(T_x)$ and let $G_x:=\bigcup_{y\in H_x}Y_y$.

% Apply \cref{generalized-tripod} to $G$ and let $\mathcal{L}$, $\mathcal{Y}$, $\mathcal{T}$, $T$, $H_x$, and $G_x$ be defined as in \cref{generalized-tripod}, where the partition $\mathcal{Y}$ has width $\ell$ with respect to the layering $\mathcal{L}$.

For a node $x\in V(T)$, we say that a shortcut $P\in\SS$ \defin{crosses} $x$ if $Y_x$ contains an internal vertex of $P$.  In other words, if $P=v_0,\ldots,v_r$ and $\{v_1,\ldots,v_{r-1}\}\cap Y_x\neq\emptyset$.  We say that a vertex $v\in V(G)$ \defin{participates} in $x$ if $v\in Y_x$, or $\SS$ contains a shortcut $P$ with $v\in V(P)$ and $P$ crosses $x$.  For each $v\in V(G)$, let $a(v)$ denote the lowest common $T$-ancestor of all the nodes $x\in V(T)$ in which $v$ participates.

% \begin{clm}
%   For each $v\in V(G)$, $v$ participates in $a(v)$.
% \end{clm}
% 
% \begin{proof}
%   Let $Z := \{v\} \cup \{\{v_1,\ldots,v_{r-1}\}:(v_0,\ldots,v_r)\in\SS, v\in \{v_0,\ldots,v_r\}\}$. Then $G[Z]$ is connected because $Z$ is the union of (vertex sets of) paths in $G$, each of which contains $v$.
% 
%   We claim that $v$ participates in a node $x\in V(T)$ if and only if $Z\cap Y_x\neq\emptyset$.  If $v$ participates in $x$ then either $v\in Y_x$, so $Z\cap Y_x\supseteq\{v\}$; or $v\in \{v_0,\ldots,v_r\}$ for some shortcut $(v_0,\ldots,v_r)\in\SS$ that crosses $x$, so $Z\cap Y_x\supseteq \{v_i\}$ for some $i\in\{1,\ldots,r-1\}$.  In the other direction, if $Z\cap Y_x\neq\emptyset$, then either $Z\cap Y_x\supseteq \{v\}$, so $v\in Y_x$; or $Z\cap Y_x\supseteq \{v_i\}$ where $i\in\{1,\ldots,r\}$, $(v_0,\ldots,v_r)\in\SS$ and $v\in\{v_0,\ldots,v_r\}$, so $v\in V(P)$ for a path $P=(v_0,\ldots,v_r)\in\SS$ that crosses $x$.
% 
%   Let $X_H:=\{x\in V(H): Z\cap Y_x\neq\emptyset\}$.  Therefore $X_H$ contains every every node in which $v$ participates.  The connectivity of $G[Z]$ implies that $H[X_H]$ is connected. The fact that $\mathcal{T}$ is a normalized tree decomposition of $H$ implies that some node $a\in X_H$ is a $T$-ancestor of every node in $X_H$.  Thus, $v$ participates in the node $a=a(v)$ that is the lowest common $T$-ancestor of every node in which $v$ participates, as required.  
% \end{proof}

% \begin{proof}
%   Let $a(v)$.  If $v\in Y_{a(v)}$ then $v$ participates in $a(v)$, by definition, and we are done.  Otherwise, there are two nodes $x,y\in V(T)$ in which $x$ participates and whose lowest common $T$-ancestor is $a(v)$.
% 
%   There are two cases to consider:  
%   \begin{compactenum}
%     \item If $v\in Y_x$ then $v\not\in Y_y$.  Since $v$ participates in $Y_y$, there exist a shortcut $P:=v_0,\ldots,v_r$ in $\SS$ with $v=v_\alpha$ and $v_\beta\in Y_y$ for some $\alpha\in\{0,\ldots,r\}$ and $\beta\in\{1,\ldots,r-1\}$.  Without loss of generality, $\alpha < \beta$.
% 
%     For each $i\in\{0,\ldots,r\}$, let $Y_{x_i}$ be the part of $\mathcal{Y}$ that contains $v_i$.  Since $G[\{v_\alpha,\ldots,v_\beta\}]$ is connected, $H[\{x_\alpha,\ldots,x_\beta\}]$ connected.  Since $\mathcal{T}$ is a normalized tree decomposition of $H$, $x_j$ is a $T$-ancestor of every node in $x_\alpha,\ldots,x_\beta$ for some $j\in\{\alpha,\ldots,\beta\}$.      In particular $x_j$ is a $T$-ancestor of $x=x_\alpha$ and $y=x_\beta$, so $x_j$ is a $T$-ancestor of $a(v)$.  Since $v_j\in V(P)$ and $v_j\in Y_{x_j}$ this implies that $v$ participates in $x_j$.  Therefore $v$ participates in a $T$-ancestor of $a(v)$, which can only mean that $v$ participates in $a(v)$.
% 
%     \item If $v\not\in Y_x$ and $v\not\in Y_y$ then there are shortcuts $P_1,P_2\in\SS$ with $v\in V(P_1)$, $v\in V(P_2)$ where $P_1$ has an internal vertex $w_1\in Y_x$ and $P_2$ has an internal vertex $w_2\in Y_y$.    
% 
% 
%     .  Since $v$ participates in $Y_y$, there exist a shortcut $v_0,\ldots,v_r$ in $\SS$ with $v=v_\alpha$ and $v_\beta\in Y_y$ for some $\alpha\in\{0,\ldots,r\}$ and $\beta\in\{1,\ldots,r-1\}$.  Without loss of generality, $\alpha < \beta$.
% 
%     For each $i\in\{0,\ldots,r\}$, let $Y_{x_i}$ be the part of $\mathcal{Y}$ that contains $v_i$.  Since $G[\{v_\alpha,\ldots,v_\beta\}]$ is connected, $H[\{x_\alpha,\ldots,x_\beta\}]$ connected.  Since $\mathcal{T}$ is a normalized tree decomposition of $H$, $x_j$ is a $T$-ancestor of every node in $x_\alpha,\ldots,x_\beta$ for some $j\in\{\alpha,\ldots,\beta\}$.  In particular $x_j$ is a $T$-ancestor of $x=x_\alpha$ and $y=x_\beta$, so $x_j$ is a $T$-ancestor of $a(v)$. Since $v_j\in Y_{x_j}$, and $v\not\in Y_{a(v)}$ this implies that $v$ participates in $Y_{x_j}$.  Therefore $v$ participates in a $T$-ancestor of $a(v)$, which means that $v$ participates in $a(v)$.
% 
% 
%   \end{compactenum}
% \end{proof}

% We let $X_v$ denote the set of nodes $x\in V(T)$ such that $v$ participates in $x$.

% \begin{clm}\label{x-v-ancestor}
%   For any $v\in V(G)$ there exists a (unique) node $a(v)\in X_v$ such that
%   $a(v)$ is a $T$-ancestor of every node in $X_v$.
% \end{clm}
% 
% \begin{proof}
%   Let $Z := \{v\} \cup \{\{v_1,\ldots,v_{r-1}\}:(v_0,\ldots,v_r)\in\SS, v\in \{v_0,\ldots,v_r\}\}$. Then $G[Z]$ is connected because $Z$ is the union of (vertex sets of) paths in $G$, each of which contains $v$.
% 
%   We claim that $v$ participates in a node $x\in V(T)$ if and only if $Z\cap Y_x\neq\emptyset$.  If $v$ participates in $x$ then either $v\in Y_x$, so $Z\cap Y_x\supseteq\{v\}$; or $v\in \{v_0,\ldots,v_r\}$ for some shortcut $(v_0,\ldots,v_r)\in\SS$ that crosses $x$, so $Z\cap Y_x\supseteq \{v_i\}$ for some $i\in\{1,\ldots,r-1\}$.  In the other direction, if $Z\cap Y_x\neq\emptyset$, then either $Z\cap Y_x\supseteq \{v\}$, so $v\in Y_x$; or $Z\cap Y_x\supseteq \{v_i\}$ where $i\in\{1,\ldots,r\}$, $(v_0,\ldots,v_r)\in\SS$ and $v\in\{v_0,\ldots,v_r\}$, so $v\in V(P)$ for a path $P=(v_0,\ldots,v_r)\in\SS$ that crosses $x$.
% 
% 
%   Let $X_H:=\{x\in V(H): Z\cap Y_x\neq\emptyset\}$.  The connectivity of $G[Z]$ implies that $H[X_H]$ is connected.  This implies that $T':=T[]
% 
% 
% 
%   %Property~\tref{ancestor-edge} implies that, for every $xy\in E(H)$, one of $x$ or $y$ is a $T$-ancestor of the other.
%   Choose $a(v)\in X_H$ to be the member of $X_H$ that does not have a strict $T$-ancestor in $X_H$.  Transitivity of the $T$-ancestor relationship, \tref{ancestor-edge}, and connectivity of $H[X_H]$ implies that such an $a(v)$ exists and is a $T$-ancestor of every node $x\in X_H$, as required.
% \end{proof}

@@

For each $x\in V(T)$, define $S_x := \{v\in V(G): a(v)= x\}$. Observe that $\PP:=(S_x : x\in V(T))$ is a sequence of pairwise disjoint sets that cover $V(G)$. Let $J:=G^\SS/\PP$ denote the resulting quotient graph. We consider $V(J)\subseteq V(T)$, where each $x\in V(J)$ is the vertex obtained by contracting $S_x$ in $G^{\SS}$. (Any node $x\in V(T)$ with $S_x=\emptyset$ does not contribute a vertex to $J$.)

@@

% First we need to understand the relationship between sets $S_x\in\PP$ and the related sets $Y_x$, $V_x$, $F_x$, and $N_x$.

% \referee{1}{8. general comment: Claims 1, 2, and 4 are especially belaboured by
% notation as they are quite straightforward.}

% \note{PM}{Same comment: Yes, these claims are definition checking.  We'll see if the experimental notation change helps.}

% \begin{clm}
% 	\label{s-subset}
%   For every $x\in V(T)$, 
%   % $S_x\subseteq V_x$.
%   $S_x\subseteq \bigcup_{x'\in V(T_x)} Y_{x'}$.
% \end{clm}
% 
% \begin{proof}
%   Let $v$ be any vertex in $S_x$ (which implies that $a(v)=x$) and let $Y_y$ be the part of $\mathcal{Y}$ that contains $v$.  By definition, $v$ participates in $y$ so by definition $a(v)$ is a $T$-ancestor of $y$, so $y\in V(T_{a(v)})$.  Therefore, $v\in Y_y\subseteq \bigcup_{x'\in V(T_x)} Y_{x'}$, as required.
% \end{proof}

% Next we complete Step~(i) and show that $\PP$ has small layered width with respect to the layering $\mathcal{L}=\langle L_1,\ldots,L_h\rangle$:

@@

% \referee{1}{10. page 10, Claim 5: Indices could mostly be removed by continuing
% to talk about $T$-ancestors instead. Move the definition of $H+$ and its
% directed version up, along with the explanation of what will be proven,
% so that it isn’t necessary to define the $s_i$.}
% 
% \note{PM}{I also found the steps in this proof tedious. Unfortunately, when I tried to give a simple explanation for it in Banff, both Piotr and Torsten remained unconvinced until we opened the paper and walked through the steps of the proof together.  I suspect the referee who can see through these proofs if one of the authors of \cite{PS21}.}
